Income Has Serious Implications On Many Facets of American Life in 2021

### 2022-03-21 outputs/paper/paper.Rmd
8th commit

All of the data used in this analysis was directly collected from the United States 2021 General Social Survey, available here: https://gss.norc.org/ (link and additional information available in the README.md file.) We downloaded the desired data in the STATA format and loaded it into R [@citeR] for manipulation. The zip file acquired included 4 files: a read me text file, the raw data as a dta file, as well as two pdf files. *GSS 2021 Codebook R1b* is essentially the instruction manual for the data set. It goes over many topics like the objectives and intents of the survey, a brief overview of the methodology and an in depth explanation of the 565 variables in the raw data. With 565 variables collected in the survey, the data file titles *gss2021.dta* has a size of 2.6 MB. Although this isn't too large of a data file, we did encounter some problems which may have implications as it pertains to reproducibility of the workflow. The following is an explanation of the problems encountered as well as an explanation of the workaround/solution we implemented to get over the obstacle.

@@

{ width="800" height="600" style="display: block; margin: 0 auto" }

@@

All the data processing, cleaning and variable selection takes place in an R script called *Data Acquisition and Processing.R* in the scripts folder. To process our data, we began by loading all the necessary libraries like haven [@citehaven] which was used to acess the information from the dta type file and we then load the data into R [@citeR]. Additionally, we used a wide variety of libraries like knitr [@citeknitr], tidyverse [@citetidyverse], tidyr [@citetidyr], janitor [@citejanitor], dplyr [@citetidyr] as well as tinytex [@citetinytex] at some points in the process for data processing and pdf document generation.

After loading the data into our local environment and saving its as a cvs called raw.csv, we began to create as subset of the data with the desired variables. There are 565 variables in the data and we obviously can't explain each of them, so here is a description of some of the variables we chose to work with. *sexnow1* is the current sex of the respondent and is a concept more similar to gender, *age*, *born*: born in America or not, *marital*: the marital status, *educ*: the highest attained level of education, *divorce*: divroced or not, whether or not they have a *degree*, *income*: their current family income, *income16*: their family income when they were 16 years old, *nathealy*: if they think enough is done for healthcare, *cohort*: year of birth, *agekdbrn*: age when first child was born and lastly, *granborn* which is a number from 0-4 signifying the number of grandparents of the respondent which were born in the united states.

Additionally, we created a new column called *income_cat* and *income16_cat* which are new variables which offer us the liberty of using the *income* and *income16* variables but as discreet, categorical variables instead of the coded numerical system used all throughout the data set and explained in the information code book titled *GSS 2021 Codebook R1b.pdf* available in the 2021_stata in the data folder in inputs.This was done by duplicating the original columns and then re-coding the values. In the case of *income* for instance, the code book explains that a value of 1 for income means the response was that the annual income was less than \$1000. As such, income_cat would read *$1,000 AND UNDER*. Similar mutations were done for the variables *sex*, *granborn*, *nathealy* and *born*.

@@

The 2021 GSS is in many ways different to the previous General Social Surveys in years prior to the ascension of the corona virus to international stardom. According to the GSS 2021 Codebook R1b, in previous iterations of the GSS, the survey was taken solely via in-person interview, where the interviewer had been adequately trained and prepared for various situations that would have arose like non responds or the person being interviewed not knowing an answer. However, for reasons related to health and safety protocols and "to safeguard the health of staff and respondents during the COVID-19 pandemic, the 2021 GSS data collection used a mail-to-web methodology instead of its traditional in-person interviews." This means that after being selected via address based sampling the selected individuals "interview" would be conducted online via a self administered questionnaire. The limitations of such will later be discussed in Section \@ref(limitations)

@@

It is important to discuss the reach of our data. We know from the provided documentation that the population group eligible for taking the survey are Adults 18 or older in the United States who live in non institutional housing at the time of interviewing. Also, the 2021 GSS  uses the last birthday method to decide which individual in the household will take the survey. This was done by having "a professional phone interviewer team conducted phone outreach to complete screeners, answer inbound phone calls from sampled households, identify respondents, and complete interviews by phone and/or prompt them to complete the web survey." As far as the invitation method goes for data collection, materials that provide a web link to invite people to participate in the survey was mailed out and an option to conduct the interview via the phone instead of a self-administered questionnaire was also provided. Out of the 27,591 lines of sample (people who had been prompted to answer the questionnaire), 4,032 responses were collected. This has implications on the finding of our analysis in Section \@ref(discussion) as we will have to consider if we are able to generalize our finding from the relatively small sample that we have to the population if all US citizens over 18 years old. The small sample size and extremely long survey are both weaknesses of the survey which we will discuss in more depth in Section \@ref(results).

@@

# Visualizing the Data and The Implications
In this analysis, I am interested at looking at how real is the American Dream and if it is found to be possible, whether or not there is any difference in attainment amoung genders. As such, in this Section \@ref(visualizing-the-data-and-the-implications), we will create various visuals which will enable us to extract as much information as we can and we will speak about the implications that they make.
```{r, echo = F, message = F, include = F}
#creating levels for the x and y axis on the graphs

@@

## Visual 1
```{r figure 1, fig.cap="write a figure caption here", echo = FALSE, fig.width=12, fig.height=10, message = F, warning = F}
income16_gender <- ggplot(mapping = aes(x = factor(income16_cat, level = level_income16), group = sex, fill = sex), data = reduced_data) +
  geom_bar(alpha = 0.7)  +
  theme_minimal() +
  labs(x = "Family Income at 16 Years Old",
       y = "Number of occurrences",
       fill = "Gender",
       title = "Total Family Income at 16 Grouped by Gender") +
  coord_flip()
income16_gender
```
In this graph (Figure \@ref(fig:figure-1)), we can proportions of family income when the responded was 16 years of age. We also see that an over-whelming majority of people in our sample had an income of over $40,000. We also see that the single largest individual group is the *NA* group. We will discuss the implications of this in Section \@ref(limitations). As we've previously discussed, when the questioned is prompted with a question, in every situation, they have the option to refuse to answer the question or say that they do not know. The both of these are listed as NA responses. We also see that of the 5 possible genders (including the NA gender, not to be confused with the NA response to income16), the overwhelming majority of respondents identified as male or female as expected.

We would like to contrast this with the current income of responds. The below is the same graph above but we've also included a similar graph but for family income at the time of response.
```{r figure2, fig.cap="write a figure caption here", echo = FALSE, fig.width=12, fig.height=10, message = F, warning = F}
#income16 vs income grouped by gender

@@

## Implications 1
In Visuals \@ref(visual-1) we can see when we look into income16 that the majority of people claimed that at the age of 16, their family income was over 40 thousand dollars. When we look over at the income variable, we are truly perplexed to see that the largest is 25 thousand or more. We will discuss the implications of the latter in Section \@ref(limitations). We can also see that the number of NA responses rises from about 530 in the case of the income at age 16 variable to approximately 900 in the income variable. This seems counter intuitive as it would appear that less people are uncertain about their family income at age 16 than they are. We suspect however that this is not true. We suspect that there could be a possible shame or embarrassment associated with sharing ones current family income as it represents the persons way of living or lifestyle today. When we speak of the American dream, we implicitly assume an increase in wealth over time and for someone who isn't as wealthy as they might want to be (likely an overwhelming large portion of the population), reluctance with sharing their current income is totally understandable. On the other hand, someones income at age 16 is psychologically less of importance to them and carries no burden in their life, as it is not an expected metric of success years down the line. We also recognize that among men and women, there is a general almost a 50:50 split at each level of income but we do notice that for income level NA, a lot of people chose not to identify their gender. It would be interesting in another analysis to analyse possible evidences of gender pay gap and how people who do not fit in the *traditional* gender roles are affected by the advent of the American Dream and whether or not their position in life hinders their ability to increse in wealth over time.

As we speak about an increase in wealth over time, we next want to see if the birth place of ones grandparents has any impact on income due to the so called *generational wealth*.

>>>>

All of the data used in this analysis was directly collected from the United States 2021 General Social Survey, available here: https://gss.norc.org/ (link and additional information available in the README.md file.) We downloaded the desired data in the STATA format and loaded it into R [@citeR] for manipulation. The zip file acquired included 4 files: a read me text file, the raw data as a dta file, as well as two pdf files. *GSS 2021 Codebook R1b* is essentially the instruction manual for the data set. It goes over many topics like the objectives and intents of the survey, a brief overview of the methodology and an in depth explanation of the 565 variables in the raw data. With 565 variables collected in the survey, the data file titled *gss2021.dta* has a size of 2.6 MB. Although this isn't too large of a data file, we did encounter some problems which may have implications as it pertains to reproducibility of the workflow. The following is an explanation of the problems encountered as well as an explanation of the workaround/solution we implemented to get over the obstacle.

@@

@@

All the data processing, cleaning and variable selection takes place in an R script called *Data Acquisition and Processing.R* in the scripts folder. To process our data, we began by loading all the necessary libraries like haven [@citehaven] which was used to acess the information from the dta type file and we then load the data into R [@citeR]. Additionally, we used a wide variety of libraries like knitr [@citeknitr], tidyverse [@citetidyverse], tidyr [@citetidyr], janitor [@citejanitor], patchwork [@citepatchwork], readr [@citereadr],dplyr [@citetidyr] as well as tinytex [@citetinytex] at some points in the process for data processing and pdf document generation.

After loading the data into our local environment and saving its as a cvs called raw.csv, we began to create subsets of the data with the desired variables. There are 565 variables in the data and we obviously can't explain each of them, so here is a description of some of the variables we chose to work with. *sexnow1* is the current sex of the respondent and is a concept more similar to gender, *age*, *born*: born in America or not, *marital*: the marital status, *educ*: the highest attained level of education, *divorce*: divroced or not, whether or not they have a *degree*, *income*: their current family income, *income16*: their family income when they were 16 years old, *nathealy*: if they think enough is done for healthcare, *cohort*: year of birth, *agekdbrn*: age when first child was born and lastly, *granborn* which is a number from 0-4 signifying the number of grandparents of the respondent which were born in the united states.

Additionally, we created new columns called *income_cat* and *income16_cat* which are new variables which offer us the liberty of using the *income* and *income16* variables as discreet, categorical variables instead of the coded numerical system used all throughout the data set and explained in the information code book titled *GSS 2021 Codebook R1b.pdf* available in the 2021_stata in the data folder in inputs. This was done by duplicating the original columns and then re-coding the values. In the case of *income* for instance, the code book explains that a value of 1 for income means the response was that the annual income was less than \$1000. As such, income_cat would read *$1,000 AND UNDER*. Similar mutations were done for the variables *sex*, *granborn*, *nathealy* and *born*.

@@

The 2021 GSS is in many ways different to the previous General Social Surveys in years prior to the ascension of the corona virus to international stardom. According to the GSS 2021 Codebook R1b, in previous iterations of the GSS, the survey was taken solely via in-person interview, where the interviewer had been adequately trained and prepared for various situations that would have arose like non responses or the person being interviewed not knowing an answer. However, for reasons related to health and safety protocols and "to safeguard the health of staff and respondents during the COVID-19 pandemic, the 2021 GSS data collection used a mail-to-web methodology instead of its traditional in-person interviews." This means that after being selected via address based sampling the selected individual's "interview" would be conducted online via a self administered questionnaire. The limitations of such will later be discussed in Section \@ref(limitations)

@@

It is important to discuss the reach of our data. We know from the provided documentation that the population group eligible for taking the survey are adults 18 or older in the United States who live in non institutional housing at the time of interviewing. Also, the 2021 GSS  uses the last birthday method to decide which individual in the household will take the survey. This was done by having "a professional phone interviewer team conducted phone outreach to complete screeners, answer inbound phone calls from sampled households, identify respondents, and complete interviews by phone and/or prompt them to complete the web survey." As far as the invitation method goes for data collection, materials that provide a web link to invite people to participate in the survey was mailed out and an option to conduct the interview via the phone instead of a self-administered questionnaire was also provided. Out of the 27,591 lines of sample (people who had been prompted to answer the questionnaire), 4,032 responses were collected. This has implications on the finding of our analysis in Section \@ref(limitations) as we will have to consider if we are able to generalize our finding from the relatively small sample that we have to the population of all US citizens over 18 years old. The small sample size and extremely long survey are both weaknesses of the survey which we will discuss in more depth in Section \@ref(limitations).

@@

# Visualizing the Data and The Implications
In this analysis, I am interested at looking at how real the American Dream is and if it is found to be possible, whether or not there is any difference in attainment amoung genders. As such, in this Section \@ref(visualizing-the-data-and-the-implications), we will create various visuals which will enable us to extract as much information as we can and we will speak about the implications that they make.
```{r, echo = F, message = F, include = F}
#creating levels for the x and y axis on the graphs

@@

## Visual 1
```{r figure 1, fig.cap="write a figure caption here", echo = FALSE, fig.width=12, fig.height=10, message = F, warning = F}
income16_gender <- ggplot(mapping = aes(x = factor(income16_cat, level = level_income16), group = sex, fill = sex), data = reduced_data) +
  geom_bar(alpha = 0.7)  +
  theme_minimal() +
  labs(x = "Family Income at 16 Years Old",
       y = "Number of occurrences",
       fill = "Gender",
       title = "Total Family Income at 16 Grouped by Gender") +
  coord_flip()
income16_gender
```
In this graph (Figure \@ref(fig:figure-1)), we can see the proportions of family income when the responded was 16 years of age. We also see that an over-whelming majority of people in our sample had an income of over $40,000. We also see that the single largest individual group is the *NA* group. We will discuss the implications of this in Section \@ref(limitations). As we've previously discussed, when the questioned is prompted with a question, in every situation, they have the option to refuse to answer the question or say that they do not know. The both of these are listed as NA responses. We also see that of the 5 possible genders (including the NA gender, not to be confused with the NA response to income16), the overwhelming majority of respondents identified as male or female as expected.

We would like to contrast this with the current income of responds. Below is the same graph as above, but we've also included a similar graph  for family income at the time of response.
```{r figure2, fig.cap="write a figure caption here", echo = FALSE, fig.width=12, fig.height=10, message = F, warning = F}
#income16 vs income grouped by gender

@@

## Implications 1
In Visuals \@ref(visual-1) we can see that when we look into income16  the majority of people claimed that at the age of 16, their family income was over 40 thousand dollars. When we look over at the income variable, we are truly perplexed to see that the largest bin is 25 thousand or more. We will discuss the implications of the latter in Section \@ref(limitations). We can also see that the number of NA responses rises from about 530 in the case of the income at age 16 variable to approximately 900 in the income variable. This seems counter intuitive as it would appear that less people are uncertain about their family income at age 16 than they are now. We suspect however that this is not true. We suspect that there could be a possible shame or embarrassment associated with sharing ones current family income as it represents the persons way of living or lifestyle today. When we speak of the American dream, we implicitly assume an increase in wealth over time and for someone who isn't as wealthy as they might want to be (likely an overwhelming large portion of the population), reluctance with sharing their current income is totally understandable. On the other hand, someones income at age 16 is psychologically less of importance to them and carries no burden in their life, as it is not an expected metric of success years down the line. We also recognize that among men and women, there is in general almost a 50:50 split at each level of income but we do notice that for income level NA, a lot of people chose not to identify their gender. It would be interesting in another analysis to analyse possible evidences of gender pay gap and how people who do not fit in the *traditional* gender roles are affected by the advent of the American Dream and whether or not their position in life hinders their ability to increase in wealth over time.

As we speak about an increase in wealth over time, we next want to see if the birth place of ones grandparents has any impact on income due to so called *generational wealth*.
